# Randomly sample PTHA18 scenarios on a source-zone
-------------------------------------------------

The PTHA18 often includes thousands or tens-of-thousands of scenarios on a
source-zone. For some applications it is impractical to work with all
scenarios, but may be practical to work with a random sample of scenarios.

This tutorial shows how to randomly sample scenarios from a given source-zone in
a manner that respects the PTHA18 scenario conditional probabilities and earthquake
magnitude-frequency modelling. 

## Get the source-zone event data
---------------------------------

The first step is to get the scenario data for the source-zone of interest.
Here we choose to work with heterogeneous-slip scenarios from the
`kermadectonga2` source-zone. 
```{r startup}
# Get the scripts to access the PTHA18
ptha18 = new.env()
source('../../get_PTHA_results.R', local=ptha18, chdir=TRUE)
# Read all heterogeneous-slip scenario metadata (slip_type='stochastic' in PTHA18)
kt2_scenarios = ptha18$get_source_zone_events_data('kermadectonga2',  slip_type='stochastic')
```

## (Optional) A quick look at the scenarios data structure

Recall that the scenario data is stored in a list, containing an `events` table,
a `unit_source_statistics` data.frame, and various filenames relevant to the source-zone.
Herein we provide a quick overview; for further discussion see [../../DETAILED_README.md](../../DETAILED_README.md).
```{r datastructure}
# Names of variables in scenario data
names(kt2_scenarios)

# Get the class of each variable
lapply(kt2_scenarios, class)
```

There are over 44-thousand scenarios in the database, with one row per scenario in the event table
```{r numscenarios}
# How many rows in the scenario events table?
nrow(kt2_scenarios$events)
```

What variables are stored in the event table? Use the `names` function to see the column names. 
```{r columnNames}
names(kt2_scenarios$events)
```

For our purposes an important variable is `Mw`, the scenario moment
magnitude. The scenario moment magnitudes are binned and take values in 7.2, 7.3, ...,
9.8. Beware that some of the higher magnitudes will be impossible according to PTHA18, and
these are assigned a rate of zero. The PTHA18 maximum-magnitude varies
depending on the source-zone, but is never greater than 9.6.
```{r mwunique}
# Print all unique Mw values -- beware not all of these will be "possible" according
# to PTHA18, as some will have zero probability of occurring.
unique(kt2_scenarios$events$Mw)
```
For the purpose of random scenario sampling, an important variable is
`rate_annual`. In general this varies for each scenario, and is equal to the scenario
conditional probability (conditional on the occurrence of an earthquake with
the same magnitude) multiplied by the logic-tree-mean-rate of scenarios with
that magnitude (events/year). 

Recall that PTHA18 uses variations in the scenario conditional probability to
account for spatial variations in tectonic convergence, to limit the scenario
peak-slip, and to adjust for bias in the earthquake-source models (although
this is more prominent for variable-area-uniform-slip scenarios than for
heterogeneous-slip scenarios). See 
[this paper](https://doi.org/10.1007/s00024-019-02299-w) for further information.

## Random scenario sampling, stratified by magnitude
----------------------------------------------------

Our simplest random scenario sampling algorithm proceeds as follows
* Group the scenarios by magnitude
* For each magnitude, sample a given number of scenarios with replacement, with the chance of sampling each scenario proportional to its conditional probability.

This can be implemented as follows (herein we select 12 scenarios for each magnitude)
```{r randomSampling1}
# Convenient shorthand for the magnitudes and rates in the event table
event_Mw = kt2_scenarios$events$Mw
event_rates = kt2_scenarios$events$rate_annual

# Make the random scenarios
random_scenarios_simple = ptha18$randomly_sample_scenarios_by_Mw_and_rate(
    event_rates=event_rates,
    event_Mw=event_Mw,
    samples_per_Mw=function(Mw){ 12 },
    mw_limits=c(7.15, 9.85))

```

The result is a `data.frame` containing the indices of the random scenarios `inds`,
their magnitudes, `mw`, as well as information on the scenario rates that will be discussed
further below.

```{r randomSampling1_B}
# Look at the first few rows
head(random_scenarios_simple)
```
The columns are
* `inds` is the indices of the randomly selected scenarios. This corresponds to indices in the `event_Mw` and `event_rates` variables. Because herein these are simply columns of the event table, `inds` also also correspond to rows in `kt2_scenarios$events`.
* `mw` is the scenario magnitude. This is the same as `event_Mw[random_scenarios_simple$inds]`
* `rate_with_this_mw` is the rate of ANY scenario with the same magnitude. Note THIS IS NOT THE RATE OF THE INDIVIDUAL SCENARIO!
* `importance_sampling_scenario_rates` is a nominal rate for each scenario which retains statistical consistency with the PTHA18. In this particular case it is simply equal to the `rate_with_this_mw` divided by the number of scenarios with that same magnitude (12 in this case). 
* `importance_sampling_scenario_rates_self_normalised` is another nominal rate for each scenario. In this case it is identical to the previous variable. However later we will consider more complex sampling methods, using importance sampling, where it may be somewhat different (basically it uses an alternative statistical estimate of the nominal scenario rate).
* `importance_sampling_scenario_weights` is equal to `importance_sampling_scenario_rates` divided by `rate_with_this_mw`. Later when we do importance sampling, this corresponds to the regular importance sampling weights. 
* `importance_sampling_scenario_weights_self_normalised` is equal to `importance_sampling_scenario_rates_self_normalised` divided by `rate_with_this_mw`. Later when we do importance sampling, this corresponds to the self-normalised importance sampling weights. 


In PTHA18 some earthquake magnitudes are impossible. In this case the scenario index will
take an `NA` value, as will various other variables. We see this at the end of the current
table, for magnitudes `9.7` and `9.8`.

```{r randomSampling1_C}
# Look at the last few rows
tail(random_scenarios_simple)
```

Aside from the impossible magnitudes, we can confirm that we have 12 scenarios per magnitude, as requested.
```{r randomSampling1_nsam}
table(random_scenarios_simple$mw)
```
