## **Installation**

It is strongly suggested that the installation is undertaken on a Linux operating 
system (e.g. the author is using Ubuntu 16.04, and the install should also work
on more recent versions). 


### **Installing R**
If you don't already have R installed, the you need to get it by following the
instructions on the [R website](https://www.r-project.org/).


### **Getting a recent version of netcdf**

You need to install the R package `ncdf4` with OPeNDAP support.  This may be
difficult unless you are running Linux (although you can do this using a
virtual machine). Non-Linux users are encouraged to install an Ubuntu virtual
machine (e.g. using [VirtualBox](https://www.virtualbox.org) or some other
virtualization software) and follow the steps below.

A further complication is that to work with the PTHA outputs on the NCI THREDDS
server you need to build the package with a reasonably recent version of
netcdf-c. This is due to a bug in netcdf-c versions prior to 4.6.1 (released
in early 2018) that prevented the remote reading of long character strings
with OPeNDAP. We store some earthquake event data as (potentially) long
character strings, and it is essential to be able to read these remotely.

At the time of writing (mid 2018), most users will have to install netcdf-c
from source to access a suitable version. Source-code for a recent release of
netcdf-c can be obtained from the [netcdf-c github
page](https://github.com/Unidata/netcdf-c/releases). You need to follow their
instructions to get it installed. While building any complex software from
source can be difficult, there is much online troubleshooting information
available, and you can ask for help on the netcdf mailing list.

Next you need to install R's `ncdf4` package, and specifically tell it to use
the newly installed netcdf.  This can be done by 
[downloading the ncdf4 sources from this site](https://cran.r-project.org/web/packages/ncdf4/index.html), 
and then running a command similar to the following, in the directory where you
have downloaded the `ncdf4` source package:

    R CMD INSTALL ncdf4_1.16.tar.gz --configure-args="--with-nc-config=/home/username/PATH_TO_YOUR_NETCDF_INSTALL/bin/nc-config"

On Ubuntu you might need to prepend `sudo` to the above command, depending on
where you do the installation. Also, you may need to adjust the numbers in the
`ncdf4_1.16.tar.gz` term above to match those of the package you download.
Furthermore, you definitely need to change the path to nc-config to match the
one on your machine.

To confirm that your `ncdf4` installation is using a suitably recent netcdf-c
library, please run the following code:
```{r checkncdf, eval=FALSE}
library(ncdf4)

# This is a file from the PTHA, describing earthquake events on the Kermadec-Tonga
# source-zone. Note I pre-pend [stringlength=4096], which prevents truncation of
# long character strings. This functionality was broken in older netcdf-c versions
test_file = paste0('[stringlength=4096]http://dapds00.nci.org.au/thredds/dodsC/fj6/',
    'PTHA/AustPTHA_1/SOURCE_ZONES/kermadectonga/TSUNAMI_EVENTS/',
    'all_stochastic_slip_earthquake_events_kermadectonga.nc')

# Open it (this will not read everything)
fid = nc_open(test_file, readunlim=FALSE)

# Try to read the event_index_string. This will be artificially truncated if 
# using an old version of netcdf-c
event_index_string = ncvar_get(fid, 'event_index_string')

#
# Report whether it worked
#

if(max(nchar(event_index_string)) == 720){
    print('Success! Your ncdf4 install can read large character strings remotely')
}else{
    print('FAIL. Perhaps ncdf4 is linking to an older netcdf-c version?')
}
```
If the above code leads to the `Success! ...` message, then the install is
working. Otherwise you will have to troubleshoot your netcdf-c install (or
your internet connection, or check for a change to the NCI THREDDS server,
etc).

### **Installing rptha**

Finally you need to install the `rptha` package. This must be built from
source, after obtaining the code from Geoscience Australia's the `PTHA` github
repository. [See instructions here](https://github.com/GeoscienceAustralia/ptha/blob/master/R/README.md)

### **Installing mapview**
Finally you should install the `mapview` package (to enable some interactive
plots shown below)
```{r installme, eval=FALSE}
install.packages('mapview')
```

## **Unit tests**
Assuming you have installed all the above dependencies, you can run the unit
tests with:
```{r unit_tests, eval=FALSE}
source('test_all.R')
```
This should print a number of 'PASS' statements, and no 'FAIL' statements. It might 
take a minute or more depending on your internet connection, because it reads datasets
from the NCI THREDDS server as part of the test.

Some of the tests will fail if you haven't installed the dependencies to read
time-series. Beware some failures can occur if your internet connection is not
performing well (or if there are issues with the NCI THREDDS server). Thus, if
the tests fail it is worth re-trying a few times to confirm it is not due to a
transient network issue.


