## **Installation**

It is strongly suggested that the installation is undertaken on a Linux
operating system. The author is using Ubuntu 18.04 (and previously used 16.04
with an updated netcdf-c build). The install should also work on more recent
versions. If you are running Windows or Mac, it will probably be easiest to use
a virtual machine (e.g. via [VirtualBox](https://www.virtualbox.org) or similar).

### **Installing R**
If you don't already have R installed, the you need to get it by following the
instructions on the [R website](https://www.r-project.org/). Use the most recent
version.

### **Installing netcdf**

You need to install the R package `ncdf4` with OPeNDAP support. This may be
difficult unless you are running Linux (although you can do this using a
virtual machine). Non-Linux users are encouraged to install an Ubuntu virtual
machine (e.g. using [VirtualBox](https://www.virtualbox.org) or some other
virtualization software). 

On recent versions of Ubuntu (18.04 and greater) you can probably install netcdf libraries like this
```
sudo apt-get install netcdf-bin libnetcdf-dev libnetcdff-dev
```
and then install netcdf4 from inside R.
```r
# Run this from inside R.
# You may need to start R with "sudo R" to get permissions to install
install.packages('ncdf4')
```

An earlier version of Ubuntu shipped netcdf with an opendap bug, which caused
problems with ptha data access. To confirm that your `ncdf4` installation is
using a suitably recent netcdf-c library, please run the following code:

```{r checkncdf, eval=FALSE}
library(ncdf4)

# This is a file from the PTHA, describing earthquake events on the Kermadec-Tonga
# source-zone. Note I pre-pend [stringlength=4096], which prevents truncation of
# long character strings. This functionality was broken in older netcdf-c versions
test_file = paste0('[stringlength=4096]http://dapds00.nci.org.au/thredds/dodsC/fj6/',
    'PTHA/AustPTHA_1/SOURCE_ZONES/kermadectonga2/TSUNAMI_EVENTS/',
    'all_stochastic_slip_earthquake_events_kermadectonga2.nc')

# Open it (this will not read everything)
fid = nc_open(test_file, readunlim=FALSE)

# Try to read the event_index_string. This will be artificially truncated if 
# using an old version of netcdf-c
event_index_string = ncvar_get(fid, 'event_index_string')

#
# Report whether it worked.
#

if(max(nchar(event_index_string)) == 756){
    print('Success! Your ncdf4 install can read large character strings remotely')
}else{
    print('FAIL. Perhaps ncdf4 is linking to an older netcdf-c version?')
}

# Shut down the connection 
nc_close(fid)
```
If the above code leads to the `Success! ...` message, then the install is
working. Otherwise you will have to troubleshoot your netcdf-c install (or
your internet connection, or check for a change to the NCI THREDDS server,
etc).

If your internet is not working perfectly, or the NCI server is down, you will see an
message like this:

```{r errorMess, eval=FALSE}
#    Error in Rsx_nc4_get_vara_double: NetCDF: DAP failure
#    Var: gaugeID  Ndims: 1   Start: 0 Count: 20185
#    Error in ncvar_get_inner(ncid2use, varid2use, nc$var[[li]]$missval, addOffset,  :  
#      C function R_nc4_get_vara_double returned error
```
In this case, just try again -- after a few attempts it usually works. If not,
then check if your internet is working. Also check whether the NCI THREDDS
server is running (occasionally it goes down for maintainence or technical
problems).


### **Installing rptha**

Finally you need to install the `rptha` package. This must be built from
source, after obtaining the code from Geoscience Australia's the `PTHA` github
repository. [See instructions here](https://github.com/GeoscienceAustralia/ptha/blob/master/R/README.md).

### **Installing mapview**
Finally you should install the `mapview` package (to enable some interactive
plots shown in the main README)
```{r installme, eval=FALSE}
install.packages('mapview')
```

## **Unit tests**
Assuming you have installed all the above dependencies, you can run the unit
tests with:
```{r unit_tests, eval=FALSE}
source('test_all.R')
```
This should print a number of 'PASS' statements, and no 'FAIL' statements. It might 
take a minute or more depending on your internet connection, because it reads datasets
from the NCI THREDDS server as part of the test.

Some of the tests will fail if you haven't installed the dependencies to read
time-series. Beware some failures can occur if your internet connection is not
performing well (or if there are issues with the NCI THREDDS server). Thus, if
the tests fail it is worth re-trying a few times to confirm it is not due to a
transient network issue.


